\plaintitle

### 2016-05-25 paper/paper.tex
Corrections and additions paper

% Load basic packages
\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead 
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}

@@

\todo{reference to sections should use the name and not a number because the numbers are not shown, so they are not usable}
In section~\ref{sec:goal} we describe the goal of the visualization and the
target audience. In section~\ref{sec:data} we describe the data used, including
its origins, advantages and limitations. In section~\ref{sec:literature} we give
an overview of related literature and web resources, including related
visualizations.  This includes both visualizations of NBA or other sports data,
as well as visualizations tackling an issue encountered during the development of
our visualization. In section~\ref{sec:visualization} we describe the
visualization itself. We give an overview of the different stages of development
of the visualization, as well as the major design decisions.
Section~\ref{sec:discussion} discusses potential improvements of the final
visualization and lessons learned from the project. We conclude in
section~\ref{sec:conclusion}. Appendix~\ref{sec:terminology} contains
definitions for common basketball terms used in this paper.

@@

The data was gathered by scraping the basketball-reference site using the
provided download capabilities. Most of the data was downloaded in csv format,
while some tables had to be manually scraped. This process was automated using
Python scripts. The data was then combined in a preprocessing step. In this
step, each team's playoff rankings were calculated based on the matches played
during the playoffs, and the rest of the data was combined into json format. The
final preprocessing step combined all data into one json file. A model of the
available data can be seen on figure~\ref{fig:data}.

\todo{Figure is still in dutch}
\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/data}
  \caption{An overview of the data.}
  \label{fig:data}
\end{figure}

@@

There are a number of traditional basketball visualizations. One of the most
popular ones is the shot chart. This visualizes the shots of a player or team
around the hoop. It provides insight into the shooting ability of a
player\cite{goldsberry,stephenchu}.

@@

England's Permier League results were visualized by Ami Sedghi for the
Guardian\ref{premierleague}. This visualization shows the team's rankings as
line graphs and allows highlighting of individual teams.

@@

The entire history of the NBA and the best teams were analyzed by
fivethirtyeight~\cite{fivethirtyeight}. They calculated an ELO-score \todo{explanation of the elo score?} for all
teams an visualized them on line graphs.

@@

\subsubsection{Context section}
The context section contains a legend for the bubble section. The bubble's
borders are gold, silver or bronze if the team ended respectively first, second
or third in the regular competition. Teams that didn't end on a podium place
during the regular competition have a red or blue stroke color indicating the
team's region, respectively the western or eastern conference. The context
section also contains a link to the ``About'' page where the user can get an
explanation of the visualization with its different views, can get some
background information about the NBA and the terminolgy used, and can check the
source of the data used in the visualization. If the user hovers over or selects 
a team in the bubble section, the context section shows more information about that 
team, as shown in Figure~\ref{fig:playoffviewteamselected}. The information shown 
includes the Simple Rating System score (SRS), the end position in the play-offs 
and the end position in the regular competition.

@@

\subsubsection{Previous designs}
Our initial design used the team circles' lightness to indicate their final
ranking in their region. The hue was used to show the league they were part of.
According to the Ranking of Perceptual Tasks~\cite{perceptualranking} this is a
good property for ordered values like the regular competition ranking. But the
combination of the logo's of the teams with the colors resulted in an overload
of information and made it more difficult to compare the end ranking of
different teams.

@@

\section{Future work}
As with most projects, the result is never 'finished', meaning that there are
always extensions and iprovements possible but due to timing or resource constraints they haven't all been implemented. We suggest a number of features that might improve the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience.
        For example when changing year, a transformation of the arrow in the
        statistics view would help the user see a difference between years.  At
        the moment the arrows change at the blink of an eye, making it difficult
        for us humans with a short term memory to make the comparison. This
        would also have been very useful in the transitions between different
        years on the bubble view.
    \item To further support our short term memory we could also 'save' the
        previous state. Eg. in the field view we could visualize ghost jerseys
        to indicate which players left the team or position and highlight
        players that joined the team.  This would help the user see how the team
        changes over time.
    \item At the moment no attention has been payed to the process, provenance
        and history of the user's exploration session. A user thus needs to
        memorize his action to be able to share or reproduce his actions. This
        could be improved by adding URL parameters. This way a user at least can
        share or save a state of an analysis. One step further is to provide a
        play-feature. The visitor could them share his exploration and replay
        it. A simple implemention would ask the visitor for a start date and an
        end date and would then 'play' the changes of a particular team and
        characteristic over time.
    \item We could support users further in looking for patterns by focussing
        more on comparisons. We could integrate a view where a visitor is able
        to compare two teams in the same year, or in two different years.
\end{itemize}

@@

\section{Lessons learned}\label{sec:discussion}
We've learned a couple of things during the creation of this visualization:
\begin{itemize}
    \item Find a story to tell with your visualization
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

\subsection{Story telling}
At the beginning of this project we were able to design the 3 separate views
quite quickly. Our main problem was to integrate these views as a whole. When we
created drafts of two views being the play-off view and the statitics view, we
suddenly noted patterns. The Golden State Warriors became NBA champion a couple
of years after a great uplift in SRS score. This was because of new players
joining the team in 2009. In the years following, the team kept on attracting
talented players and increased the inherent team score to become NBA champion in
2015. With the team view, one can even notice that the team gets stronger in
center team positions during this period. Users are able to explore the data for
patters. On top of that, we provide the insight on how the changes were impacted
by changes in teamplayers.

@@

>>>>

% Load basic packages
\usepackage{balance}       % to better equalize the last page
\usepackage{graphics}      % for EPS, load graphicx instead 
\usepackage[T1]{fontenc}   % for umlauts and other diaeresis
\usepackage{txfonts}
\usepackage{mathptmx}
\usepackage[pdflang={en-US},pdftex]{hyperref}
\usepackage{color}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{nameref}

@@

In the \nameref{sec:goal} section we describe the goal of the visualization and the
target audience. In the \nameref{sec:data} section, we describe the data used, including its origins, advantages and limitations. The \nameref{sec:literature} section gives an overview of related literature and web resources, including related
visualizations. This includes both visualizations of NBA or other sports data,
as well as visualizations tackling an issue encountered during the development of
our visualization. In the \nameref{sec:visualization} section, the visualization itself is described. We give an overview of the different stages of development
of the visualization, as well as the major design decisions.
The \nameref{sec:lessonslearned} section describes some interesting lessons we have learned from the project and in the \nameref{sec:futurework} section we discuss potential  improvements of the final visualization. We conclude in the  
\nameref{sec:conclusion} section. Appendix~\ref{sec:terminology} contains definitions 
for common basketball terms used in this paper.

@@

The data was gathered by scraping the basketball-reference site using the
provided download capabilities. Most of the data was downloaded in csv format,
while some tables had to be manually scraped. This process was automated using
Python scripts. The data was then combined in a preprocessing step. In this
step, each team's playoff rankings were calculated based on the matches played
during the playoffs, and the rest of the data was combined into json format. The
final preprocessing step combined all data into one json file. A model of the
available data can be seen on figure~\ref{fig:data} \todo{is die afbeelding nog wel juist? Is die nuttig?}.

\begin{figure}
\centering
  \includegraphics[width=1.0\columnwidth]{figures/data}
  \caption{An overview of the data.}
  \label{fig:data}
\end{figure}

@@

There are a number of traditional basketball visualizations. One of the most
popular ones is the shot chart. This visualizes the shots made by a player or team. Different methods can be used, going from a scatterplot to hexagonal charts where the shots are grouped into hexagonal regions, to a heatmap. This is often done for the the area in and around the three-point line, but it can als be done for the whole attack side of the field, as seen in the Vorped visualisation\cite{vorped}. Because shots of the own side only happen rarely, it's not shown in shot charts. Shot chart provides insight into the shooting ability of a
player\cite{goldsberry,stephenchu}.

@@

England's Permier League results were visualized by Ami Sedghi for the
Guardian\cite{premierleague}. This visualization shows the team's rankings as
line graphs and allows highlighting of individual teams.

@@

The entire history of the NBA and the best teams were analyzed by
fivethirtyeight~\cite{fivethirtyeight}. They calculated an ELO-score\cite{eloraring} 
for all teams an visualized them on line graphs.

In the article written by Johannes Becker\cite{nbaempires}, the Simple Rating System 
score (SRS score) is used to compare teams over the last 40 seasons in the NBA. The 
SRS score is represented by a color hue. For comparing quantative values, this is 
not the most ideal characteristic to use as explained by John Mackinglay\cite{automatingdesign}. Nevertheless, some interesting findings are found and presented. By giving a green stroke to the rectangle of the champion of a season, 
it's easy to see which teams have won the NBA championship frequently. This way 
they have divided the last 40 seasons in three eras being the Los Angeles Lakers 
versus the Boston Celtics era, the Michael Jordan era and the Tim Duncan era.

@@

\subsubsection{Context section}
The context section contains a legend for the bubble section. The bubble's
borders are gold, silver or bronze if the team ended respectively first, second
or third in the regular competition. Teams that didn't end on a podium place
during the regular competition have a red or blue stroke color indicating the
team's region, respectively the western or eastern conference. The context
section also contains a link to the ``About'' page where the user can get an
explanation of the visualization with its different views, can get some
background information about the NBA and the terminolgy used, and can check the
source of the data used in the visualization. If the user hovers over or selects 
a team in the bubble section, the context section shows more information about that 
team, as shown in Figure~\ref{fig:playoffviewteamselected}. The information shown 
includes the SRS score, the end position in the play-offs and the end position in 
the regular competition.

@@

\subsubsection{Previous designs}
Our initial design used the team circles' lightness to indicate their final
ranking in their region. The hue was used to show the league they were part of.
According to the Ranking of Perceptual Tasks~\cite{automatingdesign} this is a
good property for ordered values like the regular competition ranking. But the
combination of the logo's of the teams with the colors resulted in an overload
of information and made it more difficult to compare the end ranking of
different teams.

@@

@@

\section{Lessons learned}\label{sec:lessonslearned}
We've learned a couple of things during the creation of this visualization:
\begin{itemize}
    \item Find a story to tell with your visualization
    \item Each separate component should be a sufficient informative visualization on itself
    \item Creating a custom visualization costs time and opportunities
\end{itemize}
In what follows we will discuss our experience in more detail.

\subsection{Story telling}
At the beginning of this project we were able to design the 3 separate views
quite quickly. Our main concern was, when integrating these views as a whole, would 
we be able to see the relations and consequences we thought we would see and if 
we could draw the conclusions we hoped for. The relation we thought would be visible 
was that the total PER value of a team would influence the end positions in the NBA championship, both the league rank as the play-off rank. We expected that teams 
that had a big increase in total PER value, due to a good transferpolicy in a 
season, would benefit in one of the following seasons. It was not possible to draw this conclusion and according to us the reason for this is that a whole NBA team contains an average of 15 players. Some players may play a lot more than others and 
are therefore much more important to the team. This observation is why we could not draw our hoped conclusion, but at the same time we found it so interesting and we wondered if we could find an example to support this observation. The story is about 
the Chicago Bulls, the team of Michael Jordan who is seen as the greatest basketball 
player of all time. In 1984, Jordan arrives at the Chicago Bulls and in his first season he is already the top player of his team, as seen by the size of his jersey for the PER statistic. In the season 1990-1991 and the two following seasons, the 
Chicago Bulls win the NBA championship. The ``three-peat'' as it is called was done 
with almost the same core team. Next to Michael Jordan, there were around five 
players that were part of the team and were roughly the best on their position in 
the team, for example Scottie Pippen. We can see that in the seasons before this 
``three-peat'' this team was formed and already played with each other. After that 
last season form those three consecutive wins, Michael Jordan retires and the team 
is not able anymore to achieve the same. In the middle of the season 1994-1995, Michael Jordan came out of retirement. What follows is a similar story to the previous one. In the season 1995-1996 and the two following seasons, the Chicago 
Bulls again achieve to win the NBA chamionship three times in a row, again with a  team consisting of around five core players. Once again is Michael Jordan the star 
player of the team, but also Scottie Pippen is still one of the best players of the team. So to become the champion, it seems like a team needs to have a default team 
with around five core players and one or two star players.

%Our main problem was to integrate these views as a whole. When we
%created drafts of two views being the play-off view and the statitics view, we
%suddenly noted patterns. The Golden State Warriors became NBA champion a couple
%of years after a great uplift in SRS score. This was because of new players
%joining the team in 2009. In the years following, the team kept on attracting
%talented players and increased the inherent team score to become NBA champion in
%2015. With the team view, one can even notice that the team gets stronger in
%center team positions during this period. Users are able to explore the data for
%patters. On top of that, we provide the insight on how the changes were impacted
%by changes in teamplayers.

@@

\section{Future work}\label{sec:futurework}
As with most projects, the result is never 'finished', meaning that there are
always extensions and iprovements possible but due to timing or resource constraints they haven't all been implemented. We suggest a number of features that might improve the visualization:
\begin{itemize}
    \item Smooth transitions between states could improve the user experience.
        For example when changing year, a transformation of the arrow in the
        statistics view would help the user see a difference between years.  At
        the moment the arrows change at the blink of an eye, making it difficult
        for us humans with a short term memory to make the comparison. This
        would also have been very useful in the transitions between different
        years on the bubble view.
    \item To further support our short term memory we could also 'save' the
        previous state. Eg. in the field view we could visualize ghost jerseys
        to indicate which players left the team or position and highlight
        players that joined the team.  This would help the user see how the team
        changes over time.
    \item At the moment no attention has been payed to the process, provenance
        and history of the user's exploration session. A user thus needs to
        memorize his action to be able to share or reproduce his actions. This
        could be improved by adding URL parameters. This way a user at least can
        share or save a state of an analysis. One step further is to provide a
        play-feature. The visitor could them share his exploration and replay
        it. A simple implemention would ask the visitor for a start date and an
        end date and would then 'play' the changes of a particular team and
        characteristic over time.
    \item We could support users further in looking for patterns by focussing
        more on comparisons. We could integrate a view where a visitor is able
        to compare two teams in the same year, or in two different years.
    \item We could focus more on the core team, i.e. five or six players who seems 
        to be the most important for the results of a team.
\end{itemize}
